Finiteness in Cubical Type Theory

### 2020-01-17 paper.tex
roweorked beginning

  We formalise our work in Cubical Agda, and we implement a library for proof
  search (including combinators for level-polymorphic fully generic currying),
  and demonstrate how it can be used to both prove properties and synthesise
  full functions given desired properties.
\end{abstract}
\section{Introduction}
In constructive mathematics, we are often preoccupied with \emph{how} we know
something.
Finiteness is one of the prime examples which demonstrates this phenomenon: to
show that some type is finite, we might provide a bijection with another
finite set, or a surjection from some finite type.
We could alternatively say that a type is finite if any collection of its
elements larger than some bound contains duplicates, or if any stream of its
elements contains duplicates.

Classically, all of these ways to prove finiteness end up proving the same
thing: constructively, however, each defines a fundamentally different
``version'' of finiteness.
In Martin-Löf type theory \cite{martin-lofIntuitionisticTypeTheory1980}, these
variations (and more) have been explored in
\cite{spiwackConstructivelyFinite2010} and
\cite{firsovVariationsNoetherianness2016}.

Homotopy Type Theory \cite{hottbook} further adds to this story in two ways.
First, its universe of types includes non-sets: whether or not a finiteness
predicate includes these gives us another dimension on which to discriminate.

The second addition is concerned with the hiding of information.
Occasionally in a constructive setting you don't want to have to reveal how you
proved something.
To model logical disjunction, for instance, we usually use a disjoint union: the
model isn't entirely correct, though, because a proof of \(A \text{or} B\)
reveals which of \(A\) or \(B\) we actually proved.

logical disjunction, for instance, is usually modelled with a
disjoint union.
This always means, however, that to prove ``A or B'' you have to reveal which of
A or B you did prove.

The other headline feature of HoTT's generalised equality is univalence: put
simply, it allows us to treat isomorphic types as equal, giving us in the
constructive world access to a technique that is commonplace classically.
It turns out that finite types have a lot to say about isomorphisms, and we will
rely heavily on this aspect of HoTT to work with them.
\subsection{Contributions}
In this work, we will explore finite types in Cubical Type Theory
\cite{cohenCubicalTypeTheory2016}, and expose their relationship to infinite
types, species, and demonstrate their practical uses for proofs in dependently
typed programming languages.

In section~\ref{strong-finiteness}, we will explore the ``strong'' notions of
finiteness (i.e. those at least as strong as Kuratowski finiteness
\cite{kuratowskiNotionEnsembleFini1920}), with a special focus on cardinal
finiteness.
We will delineate some subtly different finiteness predicates, and prove the
relationship between each.
We will prove, then, that every discrete Kuratowski finite type is cardinal
finite, and subsequently will show that discrete Kuratowski finite sets form a
topos.

In section~\ref{species-and-containers}, we will redefine our finiteness
predicates using \emph{containers}, and prove that the new definitions are
equivalent to the old.
Using this, we will prove much stronger relationships between our finiteness
predicates and relation-based definitions of finiteness.
Finally, We will show how the species-container duality falls out naturally from
these new definitions.

In section~\ref{infinite-cardinalities}, we will extend our study of finite
types to infinite but countable types.
We will see that the finiteness predicates are mirrored with countable
counterparts, and we will prove closure under the Kleene star and plus.

Proofs of finiteness have well-known practical applications in
constructive mathematics \cite{firsovDependentlyTypedProgramming2015}.
In section~\ref{practical}, we build a library which exploits these uses in
Cubical Agda \cite{vezzosiCubicalAgdaDependently2019}, allowing automation of
complex proofs over finite types.
We frame this in terms of the principle of omniscience for finite types.
Thanks to the flexibility afforded to us by Cubical Type Theory, we are able to
go further than the usual examples of this kind of proof automation: as well as
proving properties about functions, we can synthesise functions whole-cloth from
their desired properties.
Through the unified interface for finite and countable types, we can
reuse the automation machinery for \emph{partial} proof search over infinite
search spaces.
Along the way, we extend the work in \cite{allaisGenericLevelPolymorphic2019} to
prove isomorphisms between the curried and uncurried forms of \(n\)-ary
dependent functions.

>>>>

  We formalise our work in Cubical Agda, and we implement a library for proof
  search (including combinators for level-polymorphic fully generic currying),
  and demonstrate how it can be used to both prove properties and synthesise
  full functions given desired properties.
\end{abstract}
\section{Introduction}
\subsection{Foreword}
In constructive mathematics we are often preoccupied with \emph{why} something
is true.
Take finiteness, for example.
There are a handful of ways to demonstrate some type is finite: we could provide
a surjection from another finite type; we could show that any collection of its
elements larger than some bound contains duplicates; or that any stream of its
elements contain duplicates.

Classically, all of these proofs end up proving the same thing: that our type is
finite.
Constructively (in Martin-Löf Type Theory
\cite{martin-lofIntuitionisticTypeTheory1980} at least), however, all three
of the statements above construct a different version of finiteness.
\emph{How} we show that some type is finite has a significant impact on the
type of finiteness we end up dealing with.

Homotopy Type Theory \cite{hottbook} adds another wrinkle to the story.
Firstly, in HoTT we cannot assume that every type is a (homotopy) set: this
means that the finiteness predicates above can be further split into versions
which apply to sets only, and those that apply to all types.
Secondly, HoTT gives us a principled and powerful way to construct quotients,
allowing us to regain some of the flexibility of classical definitions by
``forgetting'' the parts of a proof we would be forced to remember in MLTT.
\subsection{Contributions}
\begin{itemize}
  \item Relation between different versions of finiteness, especially card to
    bishop.
  \item Topos for Kuratowski
  \item Countable and W-topos
  \item Search
\end{itemize}
\subsection{Related Work}
\subsection{Notation}

% Taking finiteness as an example again, if we prove that some type is finite via
% bijection, in MLTT such a proof would have to reveal the particular bijection.
% This becomes problematic when defining, for instance, the category of finite
% sets: the objects of this category would ideally be types coupled with proofs of
% their finiteness, but in order to

% The five in particular that we are interested in are organised in
% figure~\ref{finite-classification}.
% We especially want to examine what separates each predicate: what extra
% information do you need, for instance, to go from a proof of Cardinal finiteness
% (section~\ref{cardinal}) to manifest Bishop finiteness
% (section~\ref{manifest-bishop}).

\input{figures/finite-classification}

% Next, we will examine the category-theoretic and topos-theoretic interpretations
% of each of the predicates.
% We will give a ``category of finite sets'' for each finiteness predicate, and
% then prove the various closure properties required to arrive at the relevant
% topos (we show that decidable Kuratowski finite sets form a \(\Pi\)-pretopos,
% for instance).

% ways to prove finiteness end up proving the same
% thing: constructively, however, each defines a fundamentally different
% ``version'' of finiteness.
% In Martin-Löf type theory \cite{martin-lofIntuitionisticTypeTheory1980}, these
% variations (and more) have been explored in
% \cite{spiwackConstructivelyFinite2010} and
% \cite{firsovVariationsNoetherianness2016}.

% Homotopy Type Theory \cite{hottbook} further adds to this story in two ways.
% First, its universe of types includes non-sets: whether or not a finiteness
% predicate includes these gives us another dimension on which to discriminate.

% The second addition is concerned with the hiding of information.
% Occasionally in a constructive setting you don't want to have to reveal how you
% proved something.
% To model logical disjunction, for instance, we usually use a disjoint union: the
% model isn't entirely correct, though, because a proof of \(A \text{or} B\)
% reveals which of \(A\) or \(B\) we actually proved.

% logical disjunction, for instance, is usually modelled with a
% disjoint union.
% This always means, however, that to prove ``A or B'' you have to reveal which of
% A or B you did prove.

% The other headline feature of HoTT's generalised equality is univalence: put
% simply, it allows us to treat isomorphic types as equal, giving us in the
% constructive world access to a technique that is commonplace classically.
% It turns out that finite types have a lot to say about isomorphisms, and we will
% rely heavily on this aspect of HoTT to work with them.
% \subsection{Contributions}
% In this work, we will explore finite types in Cubical Type Theory
% \cite{cohenCubicalTypeTheory2016}, and expose their relationship to infinite
% types, species, and demonstrate their practical uses for proofs in dependently
% typed programming languages.

% In section~\ref{strong-finiteness}, we will explore the ``strong'' notions of
% finiteness (i.e. those at least as strong as Kuratowski finiteness
% \cite{kuratowskiNotionEnsembleFini1920}), with a special focus on cardinal
% finiteness.
% We will delineate some subtly different finiteness predicates, and prove the
% relationship between each.
% We will prove, then, that every discrete Kuratowski finite type is cardinal
% finite, and subsequently will show that discrete Kuratowski finite sets form a
% topos.

% In section~\ref{species-and-containers}, we will redefine our finiteness
% predicates using \emph{containers}, and prove that the new definitions are
% equivalent to the old.
% Using this, we will prove much stronger relationships between our finiteness
% predicates and relation-based definitions of finiteness.
% Finally, We will show how the species-container duality falls out naturally from
% these new definitions.

% In section~\ref{infinite-cardinalities}, we will extend our study of finite
% types to infinite but countable types.
% We will see that the finiteness predicates are mirrored with countable
% counterparts, and we will prove closure under the Kleene star and plus.

% Proofs of finiteness have well-known practical applications in
% constructive mathematics \cite{firsovDependentlyTypedProgramming2015}.
% In section~\ref{practical}, we build a library which exploits these uses in
% Cubical Agda \cite{vezzosiCubicalAgdaDependently2019}, allowing automation of
% complex proofs over finite types.
% We frame this in terms of the principle of omniscience for finite types.
% Thanks to the flexibility afforded to us by Cubical Type Theory, we are able to
% go further than the usual examples of this kind of proof automation: as well as
% proving properties about functions, we can synthesise functions whole-cloth from
% their desired properties.
% Through the unified interface for finite and countable types, we can
% reuse the automation machinery for \emph{partial} proof search over infinite
% search spaces.
% Along the way, we extend the work in \cite{allaisGenericLevelPolymorphic2019} to
% prove isomorphisms between the curried and uncurried forms of \(n\)-ary
% dependent functions.
